Describing *Symbiodinium* metacommunities using high-throughput sequencing

### 2017-01-02 manuscript/manuscript.Rmd
incorporate Pete comments

  The composition of symbiotic algal communities (*Symbiodinium* spp.) in reef-building corals plays a major role in their biology and ecology, as the identity and functional performance of individual symbionts influences emergent outcomes such as photobiology, energetics, coral growth, and susceptibility to stress [@Rowan:2004p3383; @Cantin:2009p4376; @Jones:2011dt]. Even slight differences in the relative abundance of different symbionts within a coral may have important functional consequences [@Loram:2007p3688; @Cunning:2015ja]. Moreover, plasticity and flexibility of these communities within species and individuals over time has been identified as an important trait that may underlie a coral's sensitivity to stress [@Putnam:2012wv] and its ability to respond adaptively to environmental change by 'shuffling' or 'switching' symbionts [@Baker:2003p200]. Therefore, understanding and predicting ecological performance of corals depends on our ability to accurately and comprehensively describe their *Symbiodinium* communities.

  *Symbiodinium* are primarily described using genetic sequence data, including both chloroplast (cp23S and psbA^ncr^) and nuclear markers (18S, 28S, ITS1, and ITS2) [@LaJeunesse:2001p64; Santos:2003kt; @Pochon:2012we]. Together, these markers support nine major 'clades' within the genus *Symbiodinium* (clades A through I [@Pochon:2010p8784]), which are further subdivided into 'types' based on the marker used. *Symbiodinium* species are beginning to be described based on a combinations of markers, including microsatellites to establish reproductive isolation [@LaJeunesse:2014io]. However, until more species are described, ecological surveys of *Symbiodinium* diversity must rely on established marker genes -- the one most commonly used is ITS2. High-throughput sequencing of ITS2 is now increasingly being utilized to characterize these communities at unprecedented resolution [e.g., @Arif:2014gk; @Quigley:2014dv; @Thomas:2014da; @Edmunds:2014jc; @Cunning:2015hb]; however, these big datasets place even greater importance on issues surrounding the interpretation of ITS2 sequence diversity in relation to biological diversity.

  While ITS2 was initially investigated as a potential species-level marker in *Symbiodinium* [@LaJeunesse:2001p64], it is now understood that this marker is not variable enough in all cases [@Finney:2010p8324; @Thornhill:2013dx]; for example, the 'B1' ITS2 sequence variant is shared by *S. minutum*, *S. pseudominutum*, and *S. antillogorgium*, and potentially others [@Parkinson:2015gt]. Moreover, the position of ITS2 within the tandemly repeating ribosomal DNA array results in the occurrence of multiple ITS2 sequence variants within a single genome [@Thornhill:2007p145] that evolve through concerted evolution. In fact, this process may mask species divergence by maintaining ancestral sequence variants as numerical dominants in multiple derived lineages [@Thornhill:2013dx]. Together, these features of ITS2 complicate the interpretation of intragenomic vs. interspecific variation and preclude its use as a true species-level marker. Nevertheless, numerically dominant intragenomic variants of ITS2 are still phylogenetically informative across the *Symbiodinium* genus [@LaJeunesse:2001p64], and due to the high volume of existing sequence data for comparative analysis and the relative ease of amplifying and sequencing ITS2 using current high-throughput platforms, it remains an essential and powerful marker. Therefore, it is important to develop best practices in the bioinformatic analysis and interpretation of ITS2 metabarcoding surveys.

  Here, we describe a *Symbiodinium* metacommunity in St. John, U.S. Virgin Islands, with the objectives of 1) developing an appropriate bioinformatic approach for *Symbiodinium* ITS2 metabarcoding surveys, and 2) exploring network analysis and statistical approaches uniquely enabled by such datasets to advance understanding of *Symbiodinium* metacommunity ecology. In addressing the first objective, we test the ability of 'within-sample clustering' to extract meaningful information from ITS2 metabarcoding data given its inherent limitations. Specifically, this approach seeks to cope with the fact that sequences within a genome may be only 97% similar [@Arif:2014gk], while the dominant intragenomic variants from different species may be >97% similar, or even identical [@Thornhill:2013dx; @Parkinson:2015gt]. Clearly, treating each ITS2 sequence as a unique *Symbiodinium* type overestimates diversity due to the presence of intragenomic variants, while clustering sequences at 97% similarity underestimates diversity by collapsing species known to be different. True biological diversity may be better approximated by a within-sample clustering approach that exploits key assumptions regarding the distribution of symbionts among samples, and sequences among symbionts.

  These assumptions are that 1) most individual colonies are dominated by a single *Symbiodinium* type [@LaJeunesse:2011us; @Pettay:2011wt; @Baums:2014hj], and 2) different numerically dominant ITS2 sequence variants diagnose different *Symbiodinium* types, even when they differ by only 1 nucleotide [@Sampayo:2009p3422]. These assumptions together suggest that closely related ITS2 sequences (>97% similar; @Arif:2014gk) occurring within the same sample are more likely to arise from a single taxon rather than multiple, but closely related ITS2 sequences occurring as numerically dominant variants in different samples are more likely to represent distinct *Symbiodinium* taxa. These assumptions can be leveraged to bioinformatically parse intragenomic and interspecific variation by clustering sequences at 97% similarity within each sample independently, in order to collapse variability that is more likely to be intragenomic (i.e., occurring within a sample), while maintaining variability that is more likely to be interspecific (i.e., occurring in different samples).

  Using this within-sample clustering approach, we analyzed *Symbiodinium* communities in ten coral species across the north and south shores of St. John. We then explore ecological patterns only made possible by large metabarcoding datasets, including 1) testing for whole-community differences corresponding to environmental variability between the north and south shores, 2) analyzing coral-*Symbiodinium* metacommunity association networks, and 3) quantifying and testing for differences in symbiosis flexibility (i.e., beta diversity) among coral species. The volume of data and multitude of analysis approaches facilitated by ITS2 metabarcoding has the potential to revolutionize our understanding of *Symbiodinium* metacommunity ecology. More generally, this work demonstrates how ecological knowledge can inform bioinformatic analyses given complex marker gene to species correspondence.

@@

  A total of 87 tissue samples from scleractinian and milleporine corals were collected from various sites around St. John, US Virgin Islands, USA (Fig. 1), between August 7th and 9th, 2012. Small tissue biopsies (~5 mm diameter and ~5 mm depth) were removed from coral colonies with sterile metal borers, and the small holes were plugged with non-toxic modeling clay. Samples were preserved in ~500 µL Guanidinium buffer (50% w/v guanidinium isothiocyanate; 50 mM Tris pH 7.6; 10 μM EDTA; 4.2% w/v sarkosyl; 2.1% v/v β-mercaptoethanol) and shipped to the Hawaii Institute of Marine Biology (HIMB). Genomic DNA was extracted from each coral tissue sample following a guanidinium-based extraction protocol [@Cunning:2015hb], and sent to Research and Testing Laboratory (Lubbock, TX) for sequencing of ITS2 amplicons ('itsD' and 'its2rev2' primers from @Stat:2009p7366) on the Illumina MiSeq platform with 2x300 paired-end read chemistry.

```{r Fig1, fig.height=3, fig.width=4, fig.cap="Figure 1. Sampling locations in St. John, U.S. Virgin Islands."}
library(rgdal)
stjmap <- readOGR("data/coastline/zipfolder", layer="Coastal_Coastline_line", verbose=F)  # units are in meters
stjmap <- spTransform(stjmap, CRS("+proj=longlat +datum=WGS84"))  # transform to lat/long

@@

#pdf(file="figures/map.pdf", height=3, width=4)
par(mar=c(0,0,0,0))
plot(stjmap, lwd=1, col="gray40",
     xlim=c(-64.75, -64.68), ylim=c(18.303, 18.370))
box()
text(-64.74, 18.34, expression(italic("       St. John\nU.S. Virgin Islands")), col="gray40")
axis(side=3, at=seq(-64.75, -64.65, 0.01), line=0, tck=0.02, labels=FALSE)
axis(side=3, at=seq(-64.75, -64.65, 0.04), line=-2, lwd=0, cex.axis=0.6)
axis(side=2, at=seq(18.3, 18.4, 0.01), line=0, tck=0.02, labels=FALSE)
axis(side=2, at=seq(18.3, 18.4, 0.02), line=-2, lwd=0, cex.axis=0.6)
with(sitecoords, {
  points(lon, lat, pch=21, bg="black")
  text(lon, lat, name, pos=c(4,2,4,4,2,1,3))
})
points(viers[1], viers[2], pch=2, col="gray40", cex=0.8)
text(viers[1], viers[2], expression(italic("VIERS")), col="gray40", pos=4, cex=0.7)
#dev.off()
```

  Paired reads from each sample (provided in .fastq format by Research and Testing Laboratory) were merged using illumina-utils software [@Eren:2013hh] with an enforced Q30-check and an overlap ≥150 bp with ≤3 mismatches required to retain a sequence. Chimeric sequences were removed using usearch61 [@Edgar:2010cv] implemented in QIIME [@Caporaso:2010jf]. Primers were trimmed using cutadapt [@Martin:2011eu] allowing 3 mismatches, and only sequences with both forward and reverse primer matches and length ≥250 bp after trimming were retained. Subsequently, three different clustering approaches, each based on the uclust algorithm [@Edgar:2010cv] and implemented in QIIME, were used to group ITS2 sequences into operational taxonomic units (OTUs): 1) clustering at 100% identity, 2) clustering at 97% identity across samples (i.e., sequences from all samples clustered together), and 3) clustering at 97% identity within samples (i.e., sequences from each sample clustered independently). For each 97% cluster, the most abundant sequence variant was chosen as the representative sequence, and within-sample clusters were merged across samples if their representative sequences were 100% identical. After removing singleton clusters, representative sequences for each OTU were assigned taxonomy by searching a custom reference database of *Symbiodinium* ITS2 sequences using the Needleman-Wunsch global alignment algorithm implemented in the Biostrings package [@BiostringsStringo:Fbauxymn] in R [@RALanguageandEn:2014wf]. Each OTU was then assigned a name corresponding to the reference sequence(s) with the highest alignment score; if the match was <100%, the assignment was given a unique superscript. If the match was less than 90%, the sequence was blasted to NCBI and omitted if the top hit did not contain "Symbiodinium".

  The reference database comprised `r system("grep -c '>' data/ITS2db_trimmed_derep.fasta")` *Symbiodinium* ITS2 sequences downloaded directly from NCBI. These sequences included those used in previous reference databases [@Cunning:2015hb] supplemented with additional sequences of *Symbiodinium* from Caribbean corals [@Finney:2010p8324; @Green:2014ev]. Reference sequences were separated and aligned by clade using muscle [@Edgar:2004bo], and trimmed to equal length using the o-smart-trim command from oligotyping software [@Eren:2013fp], before being reconcatenated into a single fasta file. The bioinformatic pipeline used here was implemented using a series of bash scripts and a Makefile, which can be found in the data archive accompanying this paper and executed to fully reproduce this analysis.

  OTU tables, sample metadata, and taxonomic data were imported into R using the phyloseq package [@McMurdie:2013ky] to facilitate downstream analyses. OTU tables were filtered to remove any OTU that was not observed at least `r n` times, or any sample with less than `r sn` sequences, and counts were transformed to relative abundance. Permutational analysis of variance (PERMANOVA) was used to test for differences in *Symbiodinium* community composition between the north and south shores of St. John within each coral species.

  Network analysis and visualization was performed in R using the igraph package [@Csardi:2006vh]. For the dominant and abundant symbionts networks, weighted edges were defined as the proportion of individuals of a species in which a symbiont OTU occurred at greater than 50% or greater than 1% relative abundance, respectively. For the background symbionts network, unweighted edges were defined based on the presence of a symbiont at less than 1% relative abundance in at least one individual within the species. To simplify visualization of the background symbiont network, clade D OTUs were merged into a single node under the assumption that clade D in the Caribbean comprises a single species (*Symbiodinium trenchii*; [@LaJeunesse:2014io]), and symbiont nodes connected to ≤ 2 coral species were removed from the network. All network layouts were constructed based on the Fruchterman-Reingold algorithm.

  Beta diversity was calculated as the multivariate dispersion of samples within a coral species. Principal coordinate analysis of Bray-Curtis dissimilarities of square-root transformed OTU counts was used to calculate average distance-to-centroid values for each species, which were then compared statistically by a permutation test. This analysis was implemented using betadisper() in the vegan package [@veganCommunityEco:uw], based on [@Anderson:2006fs].

  Sea Surface Temperature (SST) and chlorophyll concentration (chla): nighttime sea surface temperature and ocean colour data at 1 km spatial resolution were extracted from the sensor MODIS aboard the Aqua satellite. We used as input the monthly climatologies available at the IMaRS website. From the 12 monthly climatological values, we calculated the median, mean, standard deviation and minimum and maximum monthly mean. Climatologies for MODIS cover the period 2003 to 2010.

  Wind-driven wave exposure for each station is dependent on the wind patterns and the configuration of the coastline, which defines the fetch, or the length of water over which a given wind has blown to generate waves. To calculate wave exposure, wind speed and direction data at each location were acquired from the QuickSCAT (NASA) satellite scatterometer from 1999 to 2008 at 25 km spatial resolution [@Hoffman:2005dz]. Coastline data were obtained from the Global Self-consistent, Hierarchical, High-resolution, Shoreline (GSHHS v 2.2) database which provides global coastline at 1:250,000 scale [@Wessel:1996wk]. From these datasets wave exposure was calculated using the methods based on wave theory described in Chollett et al. (2012) for 32 fetch directions and the coastline data at full resolution. Average wave exposure at each station was calculated in R with the aid of the packages maptools [@maptoolsToolsfor:_dYm01UK], raster [@rasterGeographicD:524kz2M0], rgeos [@rgeosInterfaceto:8WVz_nNk], and sp [@Bivand:2013ux].

@@

  The dataset comprised `r system("grep -c '>' data/fasta/combined_seqs_trimmed.fasta")` sequences after merging paired reads, removing chimeric sequences, and trimming primers. After clustering, OTUs with <`r n` sequences and samples with <`r sn` sequences were excluded, leaving n=`r nsamples(phy97.f)` samples for downstream analysis. The number of OTUs resolved, as well as the number of sequences per OTU and per sample, depended on the clustering approach used (Table 1 and Fig. S1). More OTUs were resolved as the clustering resolution increased (i.e., 97% OTUs across samples < 97% OTUs within samples < 100% OTUs), with fewer reads per OTU (Table 1). The number of sequences per sample was less for the 100% OTU approach because more OTUs were filtered out of the dataset by the minimum threshold count of `r n`.

```{r Table1}
# Compute summary statistics 
stats97.f <- data.frame(`97% OTUs<br>(across samples)`=t(phystats(phy97.f)), check.names=F)
stats97bs.f <- data.frame(`97% OTUs<br>(within samples)`=t(phystats(phy97bs.f)), check.names=F)
stats100.f <- data.frame(`100% OTUs`=t(phystats(phy100.f)), check.names=F)

@@

  A subset of samples were selected for comparative analysis of the three clustering approaches (Fig. 1). 97% across-sample clustering assigned the same dominant OTU to samples dominated by different sequence variants, while 97% within-sample clustering assigned distinct dominant OTUs to these samples. Because the outcomes of this approach are more consistent with the current understanding of ITS2 sequence diversity as it links to *Symbiodinium* biology and ecology (see Discussion), we present the remainder of the results using the 97% within-sample clustering approach. Additional detailed comparisons of the outcomes of each clustering approach for each coral species are presented in the Supplementary Information.

```{r Fig2, fig.width=6, fig.height=4, fig.align="center", fig.cap="Figure 2. Comparison of dominant OTUs assigned by 100%, 97% across-sample, and 97% within-sample clustering approaches. A.) The composition of unique sequence variants (i.e., 100% OTUs) in each sample, sorted by relative abundance, and filled by unique colors corresponding to unique sequence variants. The dominant OTUs within each sample assigned by B.) 97% across-sample clustering and C.) 97% within-sample clustering are shown by rectangles below each bar, with fill colors matching the unique sequence variants presented in (A.) to indicate the representative sequence of the assigned OTU. Colored rectangles that span multiple bars indicate that the corresponding samples were assigned the same dominant OTU. Rectangles are annotated with the closest BLAST hit to the representative sequence of the OTU; those with the same annotation but different fill colors represent unique sequence variants with the same closest BLAST hit."}
set.seed(42432)
otubarplot2(samples=c("64","72","29","45","73","59","47","26","48","66","63"))
```

@@

#### Differences between shores

>>>>

**Be very explicit up front about what is meant by type vs species**

  The composition of symbiotic algal communities (*Symbiodinium* spp.) in reef-building corals plays a major role in their biology and ecology, as the identity and functional performance of individual symbionts influences emergent properties of the holobiont, including its photobiology, energetics, growth rates, and susceptibility to stress [@Rowan:2004p3383; @Cantin:2009p4376; @Jones:2011dt]. Even slight differences in the relative abundance of different symbionts within a coral may have important functional consequences [@Loram:2007p3688; @Cunning:2015ja]. Moreover, plasticity and flexibility of these communities within species and individuals over time is an important trait modulating the sensitivity of corals to environmental stress [@Putnam:2012wv] and their ability to respond adaptively by 'shuffling' or 'switching' symbionts [@Baker:2003p200]. Therefore, understanding the ecological performance of corals depends on accurately and comprehensively describing their *Symbiodinium* communities.

  *Symbiodinium* are primarily described using genetic sequence data, including both chloroplast (cp23S and psbA^ncr^) and nuclear markers (18S, 28S, ITS1, and ITS2) [@LaJeunesse:2001p64; Santos:2003kt; @Pochon:2012we]. Together, these markers support nine major 'clades' within the genus *Symbiodinium* (clades A through I [@Pochon:2010p8784]), which are further subdivided into 'types' based on the marker used to identify them. *Symbiodinium* species are beginning to be described based on a combinations of markers, including microsatellites to establish reproductive isolation (i.e., to satisfy the biological species concept) [@LaJeunesse:2014io]. However, until more species are described, ecological surveys of *Symbiodinium* diversity must rely on established marker genes, of which the most commonly used is ITS2. High-throughput sequencing of ITS2 increasingly is being utilized to characterize these communities with unprecedented resolution [e.g., @Arif:2014gk; @Quigley:2014dv; @Thomas:2014da; @Edmunds:2014jc; @Cunning:2015hb]. However, such analyses create extremely large datasets consisting of millions of sequences, which places great importance on the ways in which ITS2 sequence diversity is analyzed and interpreted in relation to biological diversity.

  While ITS2 initially was investigated as a potential species-level marker in *Symbiodinium* [@LaJeunesse:2001p64], it is now understood that this marker is not sufficiently variable to distinguish all species within this genus [@Finney:2010p8324; @Thornhill:2013dx]. For example, the 'B1' ITS2 sequence variant is shared by *S. minutum*, *S. pseudominutum*, and *S. antillogorgium*, and potentially other species [@Parkinson:2015gt]. Moreover, the position of ITS2 within the tandemly repeating ribosomal DNA array results in the occurrence of multiple ITS2 sequence variants within a single genome [@Thornhill:2007p145] that evolve through concerted evolution (sensu XXX). In fact, concerted evolution may mask species divergence by maintaining ancestral sequence variants as numerical dominants in multiple derived lineages [@Thornhill:2013dx]. Together, these features of ITS2 complicate the interpretation of intragenomic versus interspecific variation, and preclude its use as a true species-level marker. Nevertheless, numerically dominant intragenomic variants of ITS2 are still phylogenetically informative across the *Symbiodinium* genus [@LaJeunesse:2001p64], and due to the high volume of existing sequence data for comparative analysis (e.g., GeoSymbio, etc.), and the relative ease of amplifying and sequencing ITS2 using current high-throughput platforms (XXX), it remains an essential and powerful marker for the *Symbiodinium* system. Therefore, it is important to develop best practices in the bioinformatic analysis and interpretation of ITS2 metabarcoding surveys.

  Here, we describe a *Symbiodinium* metacommunity with scleractinians and a *Millepora* hydrocoral in St. John, U.S. Virgin Islands, with the objectives of: 1) developing an appropriate bioinformatic approach for *Symbiodinium* ITS2 metabarcoding surveys, and 2) exploring network analysis and statistical applications by which such datasets may advance our understanding of *Symbiodinium* metacommunity ecology. In addressing the first objective, we test the ability of 'within-sample clustering' to extract ecologically relevant information from ITS2 metabarcoding data given its aforementioned limitations. Specifically, within-sample clustering addresses the fact that sequences within one species' genome may be only 97% similar [@Arif:2014gk], while the dominant intragenomic variants from different species may be >97% similar, or even identical [@Thornhill:2013dx; @Parkinson:2015gt]. Clearly, treating each ITS2 sequence as a unique *Symbiodinium* type overestimates species diversity due to the presence of intragenomic variants, while clustering sequences at 97% similarity underestimates species diversity by collapsing species known to be different. True biological diversity in *Symbiodinium* may be better approximated by a within-sample clustering approach that exploits key assumptions regarding the distribution of symbionts among samples, and sequences among symbionts. These assumptions include: 1) that most coral colonies are dominated by a single *Symbiodinium* type [@LaJeunesse:2011us; @Pettay:2011wt; @Baums:2014hj], and 2) that different numerically dominant ITS2 sequence variants diagnose different *Symbiodinium* types, even when they differ by only 1 nucleotide [@Sampayo:2009p3422]. These assumptions suggest that when closely related ITS2 sequences (i.e., that are >97% similar; @Arif:2014gk) occur within the same sample, they are more likely to represent a single taxon rather than multiple, but when they are numerically dominant in different samples, they are more likely to represent distinct *Symbiodinium* taxa. Accordingly, clustering sequences at 97% similarity within each sample independently may collapse variability that is more likely to be intragenomic (i.e., occurring within a sample), while maintaining variability that is more likely to be interspecific (i.e., occurring in different samples).

  Using this within-sample clustering approach, we analyzed *Symbiodinium* communities in ten coral species across the north and south shores of St. John. We explore ecological patterns that can be revealed by large-scale metabarcoding datasets, including 1) testing for whole-community differences associated with north and south shore locations, 2) analyzing coral-*Symbiodinium* metacommunity association networks, and 3) testing for differences in symbiosis flexibility (i.e., beta diversity) among coral species. The large quantity of data and the kinds of analytical approaches facilitated by ITS2 metabarcoding have the potential to revolutionize our understanding of *Symbiodinium* metacommunity ecology in reef corals and other taxa harboring similar symbionts. More generally, this work demonstrates how ecological knowledge can inform bioinformatic analyses given complex marker gene to species correspondence.

@@

  A total of 87 tissue samples from scleractinian and milleporine corals were collected from seven sites around St. John, US Virgin Islands, USA (Fig. 1), between August 7th and 9th, 2012. Samples were collected on snorkel from 3-7m depth, and all colonies were selected haphazardly in upward facing locations exposed to ambient irradiance at the sampling depth. Tissue biopsies (~5 mm diameter and ~5 mm depth) were removed from coral colonies with sterile metal borers, and the small holes were plugged with non-toxic modeling clay. Samples were preserved in ~500 µL Guanidinium buffer (50% w/v guanidinium isothiocyanate; 50 mM Tris pH 7.6; 10 μM EDTA; 4.2% w/v sarkosyl; 2.1% v/v β-mercaptoethanol) and shipped to the Hawaii Institute of Marine Biology (HIMB). Genomic DNA was extracted from each coral tissue sample following a guanidinium-based extraction protocol [@Cunning:2015hb], and sent to Research and Testing Laboratory (Lubbock, TX) for sequencing of ITS2 amplicons ('itsD' and 'its2rev2' primers from @Stat:2009p7366) on the Illumina MiSeq platform with 2x300 paired-end read chemistry.

  Broad-scale differences in the hydrographic regimes of the north and south shores of St. John were evaluated from remote sensing tools to determine sea surface temperature (°C), surface chlorophyll concentrations (mg chl a m^-3^), and wave regime. These features were selected as they corresponded to anecdotal observations of differences between the water masses on both shores, and reflected conditions that strongly affect coral performance and *Symbiodinium* biology (i.e., SST [xxx] and wave regime [xxx]), and reflect nutrients and planktonic food resources that play important roles in coral nutrition (xxx). SST was evaluated from nighttime records, and chlorophyll from ocean colour data, both of which were obtained at 1 km spatial resolution from the MODIS sensor aboard the Aqua satellite [xxx]. From the 12 monthly climatological values obtained for each sampling year (2003-2010, n=96 months), the median, mean, standard deviation and minimum and maximum monthly mean were calculated.

  Wind-driven wave exposure for a given site is dependent on the wind patterns and the configuration of the coastline, which defines the fetch, or the length of water over which a given wind has blown to generate waves. To calculate wave exposure, wind speed and direction data at each location were acquired from the QuickSCAT (NASA) satellite scatterometer from 1999 to 2008 at 25 km spatial resolution [@Hoffman:2005dz]. Coastline data were obtained from the Global Self-consistent, Hierarchical, High-resolution, Shoreline (GSHHS v 2.2) database, which provides global coastline at 1:250,000 scale [@Wessel:1996wk]. From these data, wave exposure was calculated using the methods based on wave theory (after Chollett et al. 2012) for 32 fetch directions and coastline data at full resolution. Mean wave exposure (J m^-3^) at each station was calculated in R using the packages maptools [@maptoolsToolsfor:_dYm01UK], raster [@rasterGeographicD:524kz2M0], rgeos [@rgeosInterfaceto:8WVz_nNk], and sp [@Bivand:2013ux].

  Paired reads from each sample (provided in .fastq format by Research and Testing Laboratory) were merged using illumina-utils software [@Eren:2013hh] with an enforced Q30-check and an overlap ≥150 bp with ≤3 mismatches required to retain a sequence. Chimeric sequences were removed using usearch61 [@Edgar:2010cv] implemented in QIIME [@Caporaso:2010jf]. Primers were trimmed using cutadapt [@Martin:2011eu] allowing 3 mismatches, and only sequences with both forward and reverse primer matches and length ≥250 bp after trimming were retained. Subsequently, three different clustering approaches, each based on the uclust algorithm [@Edgar:2010cv] and implemented in QIIME, were used to group ITS2 sequences into operational taxonomic units (OTUs): 1) clustering at 100% identity, 2) clustering at 97% identity across samples (i.e., sequences from all samples clustered together), and 3) clustering at 97% identity within samples (i.e., sequences from each sample clustered independently). For each 97% cluster, the most abundant sequence variant was chosen as the representative sequence, and within-sample clusters were merged across samples if their representative sequences were 100% identical. After removing singleton clusters, representative sequences for each OTU were assigned taxonomy by searching a custom reference database of *Symbiodinium* ITS2 sequences using the Needleman-Wunsch global alignment algorithm implemented in the Biostrings package [@BiostringsStringo:Fbauxymn] in R [@RALanguageandEn:2014wf]. Each OTU was then assigned a name corresponding to the reference sequence(s) with the highest alignment score; if the match was <100%, the assignment was given a unique superscript. If the match was <90%, the sequence was blasted to NCBI and omitted if the top hit did not contain "Symbiodinium".

  The reference database comprised `r system("grep -c '>' data/ITS2db_trimmed_derep.fasta")` *Symbiodinium* ITS2 sequences downloaded directly from NCBI. These sequences included those used in previous reference databases [@Cunning:2015hb] supplemented with additional sequences of *Symbiodinium* from Caribbean corals [@Finney:2010p8324; @Green:2014ev]. Reference sequences were separated and aligned by clade using muscle [@Edgar:2004bo], and trimmed to equal length using the o-smart-trim command from oligotyping software [@Eren:2013fp], before being reconcatenated into a single fasta file. The bioinformatic pipeline used here was implemented using a series of bash scripts and a Makefile, which can be found in the data archive accompanying this paper and executed to fully reproduce the present analysis.

  OTU tables, sample metadata, and taxonomic data were imported into R using the phyloseq package [@McMurdie:2013ky] to facilitate downstream analyses. OTU tables were filtered to remove any OTU that was not observed at least `r n` times, or any sample with < `r sn` sequences, and counts were transformed to relative abundance. Permutational analysis of variance (PERMANOVA) was used to test for differences in *Symbiodinium* community composition between the north and south shores of St. John within each coral species.

  Network analysis and visualization was performed in R using the igraph package [@Csardi:2006vh]. Networks were created featuring 'dominant' (>50% relative abundance) and 'abundant' (>1% relative abundance) OTUs, with weighted edges proportional to the number of individuals within a species in which a symbiont OTU occurred. A 'background' symbionts network was also created with unweighted edges defined based on the presence of a symbiont at <1% relative abundance in at least one individual within the species. To simplify visualization of the background symbiont network, clade D OTUs were merged into a single node under the assumption that clade D in the Caribbean comprises a single species (*Symbiodinium trenchii*; [@LaJeunesse:2014io]), and symbiont nodes connected to ≤ 2 coral species were removed from the network. All network layouts were constructed based on the Fruchterman-Reingold algorithm (XXX).

  Beta diversity (sensu XXX) was calculated as the multivariate dispersion of samples within a coral species. Principal coordinate analysis of Bray-Curtis dissimilarities of square-root transformed OTU counts was used to calculate average distance-to-centroid values for each species, which were then compared statistically by a permutation test. This analysis was implemented using betadisper() in the vegan package [@veganCommunityEco:uw], based on [@Anderson:2006fs].

#### Environmental and sequence data overview

  The environmental conditions broadly characterizing the north and south shores are presented in Figure 1. From 2003 to 2010, the north shore of St. John was characterized by lower wave exposure, slightly higher and more variable sea surface temperatures, and higher and more variable chlorophyll a concentrations, relative to the south shore. (Statistical tests?)

```{r Fig1, fig.height=3, fig.width=5, fig.cap="Figure 1. Sampling locations around St. John, U.S. Virgin Islands."}
library(rgdal)
stjmap <- readOGR("data/coastline/zipfolder", layer="Coastal_Coastline_line", verbose=F)  # units are in meters
stjmap <- spTransform(stjmap, CRS("+proj=longlat +datum=WGS84"))  # transform to lat/long

@@

#pdf(file="figures/map.pdf", height=3, width=6)
layout(mat=matrix(c(1,1,1,2,1,1,1,3,1,1,1,4), nrow=3, byrow=T))
par(mar=c(0,0,0,0), oma=c(0,0,0,0), cex=1)
plot(stjmap, lwd=1, col="gray40",
     xlim=c(-64.74, -64.68), ylim=c(18.303, 18.370))
#box()
text(-64.735, 18.34, expression(italic("       St. John\nU.S. Virgin Islands")), col="gray40")
axis(side=3, at=seq(-64.77, -64.66, 0.01), line=0, tck=0.02, labels=FALSE)
axis(side=3, at=seq(-64.77, -64.66, 0.04), line=-2, lwd=0, cex.axis=0.6)
axis(side=2, at=seq(18.3, 18.4, 0.01), line=0, tck=0.02, labels=FALSE)
axis(side=2, at=seq(18.3, 18.4, 0.02), line=-2, lwd=0, cex.axis=0.6)
with(sitecoords, {
  points(lon, lat, pch=21, bg="black")
  text(lon, lat, name, pos=c(4,2,4,4,2,1,3))
})
points(viers[1], viers[2], pch=2, col="gray40", cex=0.8)
text(viers[1], viers[2], expression(italic("VIERS")), col="gray40", pos=4, cex=0.7)
NS <- c(-64.711967, 18.368141)
SS <- c(-64.726840, 18.315465)
points(NS[1], NS[2], pch="*", cex=2, col="red")
points(SS[1], SS[2], pch="*", cex=2, col="red")
#dev.off()

# Plot environmental data
envdata <- read.csv("data/STJ_envdata.csv", check.names=F, row.names=1)
sites <- envdata[-1,4:5]
colnames(sites) <- c("North", "South")
sites$North <- as.numeric(as.character(sites$North))
sites$South <- as.numeric(as.character(sites$South))
sites <- data.frame(t(sites))

#SST
sst <- sites[, 2:6]
sst <- within(sst, {
  `Mean+S.D.` <- Mean.SST + S.D..SST
  `Mean-S.D.` <- Mean.SST - S.D..SST
})

par(cex=0.7, mar=c(2,0,1,2), mgp=c(1.5,0.5,0), xpd=NA)
plot(NA, xlim=c(26,29), ylim=c(0.5,2.5), bty="n", yaxt="n", ylab="", xlab="SST (°C)", xaxt="n")
axis(side=1, at=seq(26,29,1))
bxp(list(stats=matrix(as.numeric(sst[1, c(1,7,3,6,2)],5,1)), n=20), add=T, at=2, axes=F, horiz=T)
bxp(list(stats=matrix(as.numeric(sst[2, c(1,7,3,6,2)],5,1)), n=20), add=T, at=1, axes=F, horiz=T)

#chlA
chla <- sites[, 7:11]
chla <- within(chla, {
  `Mean+S.D.` <- Mean.Chla + S.D..Chla
  `Mean-S.D.` <- Mean.Chla - S.D..Chla
})

#par(new=T, mar=c(7,14,10,8), cex=0.7)
plot(NA, xlim=c(0.03,0.09), ylim=c(0.5,2.5), bty="n", yaxt="n", ylab="", xlab="ChlA (mg m-3)", xaxt="n")
axis(side=1, at=seq(0.03,0.09,0.01))
bxp(list(stats=matrix(as.numeric(chla[1, c(1,7,3,6,2)],5,1)), n=20), add=T, at=2, axes=F, horiz=T)
bxp(list(stats=matrix(as.numeric(chla[2, c(1,7,3,6,2)],5,1)), n=20), add=T, at=1, axes=F, horiz=T)

t(nsamp)[2,]
#dev.off()
```

  Sequence data were successfully generated from 84 coral samples collected across two sites on the north shore and five sites on the south shore of St. John (Fig. 1). The dataset comprised `r system("grep -c '>' data/fasta/combined_seqs_trimmed.fasta")` sequences after merging paired reads, removing chimeric sequences, and trimming primers. After clustering, OTUs with <`r n` sequences and samples with <`r sn` sequences were excluded, leaving `r nsamples(phy97.f)` samples for downstream analysis.

@@

   The number of OTUs resolved, as well as the number of sequences per OTU and per sample, depended on the clustering approach used (Table 1 and Fig. S1). More OTUs were resolved as the clustering resolution increased (i.e., 97% OTUs across samples < 97% OTUs within samples < 100% OTUs), with fewer reads per OTU (Table 1). The number of sequences per sample was less for the 100% OTU approach because more OTUs were filtered out of the dataset by the minimum threshold count of `r n`.

```{r Table2}
# Compute summary statistics 
stats97.f <- data.frame(`97% OTUs<br>(across samples)`=t(phystats(phy97.f)), check.names=F)
stats97bs.f <- data.frame(`97% OTUs<br>(within samples)`=t(phystats(phy97bs.f)), check.names=F)
stats100.f <- data.frame(`100% OTUs`=t(phystats(phy100.f)), check.names=F)

@@

  A subset of samples was selected for comparative analysis of the three clustering approaches (Fig. 1). 97% across-sample clustering assigned the same dominant OTU to samples dominated by different sequence variants, while 97% within-sample clustering assigned distinct dominant OTUs to these samples. Because the outcomes of this approach are more consistent with the current understanding of ITS2 sequence diversity as it links to *Symbiodinium* biology and ecology (see Discussion), the remainder of the results is presented using the 97% within-sample clustering approach. Additional detailed comparisons of the outcomes of each clustering approach for each coral species are presented in the Supplementary Information.

```{r Fig2, fig.width=6, fig.height=4, fig.align="center", fig.cap="Figure 2. Comparison of dominant OTUs assigned by 100%, 97% across-sample, and 97% within-sample clustering approaches. A.) The composition of unique sequence variants (i.e., 100% OTUs) in each sample, sorted by relative abundance, and filled by unique colors corresponding to unique sequence variants. The dominant OTUs within each sample assigned by B.) 97% across-sample clustering and C.) 97% within-sample clustering are shown by rectangles below each bar, with fill colors matching the unique sequence variants presented in (A.) to indicate the representative sequence of the assigned OTU. Colored rectangles that span multiple bars indicate that the corresponding samples were assigned the same dominant OTU. Rectangles are annotated with the closest BLAST hit to the representative sequence of the OTU; those with the same annotation but different fill colors represent unique sequence variants with the same closest BLAST hit."}
set.seed(42432)
par(mfrow=c(1,1))
otubarplot2(samples=c("64","72","29","45","73","59","47","26","48","66","63"))
```

@@

#### Differences in *Symbiodinium* between shores
